---
title: "Chat with the Corpus"
format:
  html:
    theme: [flatly, assets/custom.scss]
    page-layout: custom
server: shiny
---

Retrieval-augmented Q&A over the same primary-source corpus used in the
[continuity/rupture analysis](index.qmd) — Catholic, Anglican, Orthodox,
Southern Baptist, and USCCB statements from 1845–2026. Gemini searches the
1,474-document corpus itself via a tool call before answering, and cites
every claim by tradition and year.

This is a live [Shiny](https://shiny.posit.co/) document (built with
[shinychat](https://posit-dev.github.io/shinychat/) and
[ellmer](https://ellmer.tidyverse.org/)) — it needs a running R process, so
it's hosted separately from the rest of this static site; see
[rag/README.md](rag/README.md) for deployment.

```{r}
#| context: setup
library(shiny)
library(bslib)
library(shinychat)
library(ellmer)
library(jsonlite)

corpus <- fromJSON(file.path("rag", "corpus.json"))

STOPWORDS <- c(
  "the", "and", "for", "are", "was", "were", "that", "this", "with", "from",
  "have", "has", "had", "not", "but", "you", "your", "our", "his", "her",
  "its", "who", "what", "when", "where", "why", "how", "did", "does", "about"
)

# Plain keyword scoring, not embeddings — Claude calls this as a tool when it
# needs evidence, rather than the app pre-retrieving before every turn.
search_corpus <- function(query, tradition = "") {
  df <- corpus
  if (nzchar(tradition)) {
    df <- df[df$tradition == tradition, , drop = FALSE]
  }
  if (nrow(df) == 0) {
    return("No documents found for that tradition.")
  }

  terms <- unique(tolower(strsplit(query, "\\s+")[[1]]))
  terms <- terms[nchar(terms) > 3 & !(terms %in% STOPWORDS)]
  if (length(terms) == 0) {
    return("Query too short to search on.")
  }

  text_lower <- tolower(df$text)
  scores <- vapply(text_lower, function(doc_text) {
    sum(vapply(terms, function(term) {
      lengths(regmatches(doc_text, gregexpr(term, doc_text, fixed = TRUE)))
    }, integer(1)))
  }, numeric(1))

  top_idx <- order(scores, decreasing = TRUE)[seq_len(min(6, nrow(df)))]
  top_idx <- top_idx[scores[top_idx] > 0]
  if (length(top_idx) == 0) {
    return("No matching passages found in the corpus for that query.")
  }

  excerpts <- df[top_idx, , drop = FALSE]
  paste(
    sprintf(
      '<excerpt tradition="%s" year="%s" title="%s">\n%s\n</excerpt>',
      excerpts$tradition,
      excerpts$year,
      excerpts$title,
      substr(excerpts$text, 1, 3000)
    ),
    collapse = "\n\n"
  )
}
```

```{r}
bslib::page_fillable(
  chat_ui(
    "chat",
    placeholder = "e.g. How did Catholic and Anglican statements on religious liberty diverge after 1960?"
  )
)
```

```{r}
#| context: server
system_prompt <- paste(
  "You are a research assistant answering questions about continuity and",
  "divergence across Christian traditions (Catholic, Anglican, Orthodox,",
  "Southern Baptist, and USCCB statements), based on a corpus of",
  "denominational encyclicals, resolutions, and conference statements",
  "spanning the 19th-21st centuries.",
  "",
  "Before answering, call the search_corpus tool to find relevant excerpts",
  "-- do not answer from general knowledge. Cite every claim inline like",
  "(Catholic, 1965) using the excerpts' tradition and year tags. If",
  "search_corpus returns nothing relevant, say so explicitly rather than",
  "guessing.",
  sep = "\n"
)

chat_client <- chat_google_gemini(
  system_prompt = system_prompt,
  model = "gemini-2.5-flash"
)

chat_client$register_tool(tool(
  search_corpus,
  description = paste(
    "Search the denominational-statement corpus for passages relevant to a",
    "query. Returns up to 6 excerpts tagged with tradition, year, and title."
  ),
  arguments = list(
    query = type_string("The search query -- keywords or a paraphrased question."),
    tradition = type_string(paste(
      "Optional: restrict results to one tradition. One of catholic,",
      "anglican, orthodox, sbc, usccb, or an empty string to search all."
    ))
  )
))

observeEvent(input$chat_user_input, {
  stream <- chat_client$stream_async(input$chat_user_input)
  chat_append("chat", stream)
})
```
